---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NaturalSpline

<!-- badges: start -->
<!-- badges: end -->

# Overview

This package provides tools for performing natural spline polynomial extrapolations on the boundary points of a given signal and the filtered signal. In methods like **trend filtering**, boundary points often fail to converge well at the two ends. This package addresses the issue by replacing the boundary values with natural spline polynomial extrapolations.

The package includes functions to compute extrapolation matrices and perform compact and non-compact extrapolations. It is especially useful for denoising applications where **trend filtering** results in inaccurate boundary estimates.

---

# Core Functions

## Extrapolation Functions

### `extrapolate_compact(x, theta, m, side)`
Performs natural spline extrapolation in a **compact** form using a formula-based approach.

- **Inputs:**
  - `x`: A numeric vector of signal values (ordered).
  - `theta`: A numeric vector of denoised values corresponding to `x`.
  - `m`: Degree of the polynomial extrapolation.
  - `side`: `"left"` or `"right"` indicating which boundary to extrapolate.
- **Output:**
  - A vector of extrapolated values.

### `extrapolate_noncompact(x, theta, m, side)`
Uses matrix multiplication for extrapolation. The matrices are precomputed from helper functions.

- **Inputs and outputs:** Same as `extrapolate_compact`.

---

## Matrix Construction Functions

### `compute_P_matrix(x, m, side)`
Computes the polynomial \( P \) matrix based on the target and support points for the specified boundary (`"left"` or `"right"`).

### `compute_C_matrix_withoutTheta(x, m, side)`
Computes the coefficient \( C \) matrix (without \(\theta\)) using the natural spline formula.

### `compute_A_matrix(P, C_withoutTheta)`
Combines the \( P \) matrix and \( C \) matrix (without \(\theta\)) to compute the \( A \) matrix, which will be multiplied with the \(\theta\)) vector.

---

## Visualization

### `plot_extrapolation(x, theta, left_result, right_result)`
Creates a plot comparing the original signal (`x`), the denoised signal (`theta`), and the extrapolated values at the boundary.

- **Features:**
  - Plots `x` (blue line), `theta` (red line), and extrapolated points (green line for the boundary).

---

# Usage Example

```{r example}
library(NaturalSpline)

# Input signal
x <- c( 5, 5.5, 6, 4, 5, 7, 8)
theta <- c( 4, 5, 5, 4.5, 5, 3, 2)
m <- 3

# Perform extrapolation using the compact formula
left_result <- extrapolate_compact(x, theta, m, side = "left")
right_result <- extrapolate_compact(x, theta, m, side = "right")
print(left_result)
print(right_result)

# Visualize results
plot_extrapolation(x, theta, m)

```

## Installation

You can install the development version of NaturalSpline from [GitHub](https://github.com/zhoumo2716/NaturalSpline) with:

```{r}
# install.packages("pak")
pak::pak("zhoumo2716/NaturalSpline")
```

Or with `devtools`
```{r}
# Install the devtools package if not already installed
install.packages("devtools")

# Install the NaturalSpline package from GitHub
devtools::install_github("zhoumo2716/NaturalSpline")
```
